---
title: "Example Analysis"
author: "Peter Freeman"
date: "June 30, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following chunk contains an implementation of the "honest tree" approach of Wager et al. (2014) and Wager & Athey (2015).
This chunk also appears in RF_TST.Rmd. It may be considered a black box.
```{r}
library(partykit)
library(rpart)

rf_prediction = function(rf,rf.data,pred.data=rf.data,prog.bar=FALSE)
{
  if ( is.null(rf$inbag) ) stop("Random forest must be trained with keep.inbag = TRUE")
  if ( length(unique(colSums(rf$inbag))) > 1 ) {
    stop("The keep.inbag field must store the number of times each observation was used.
         \nMake sure the latest version of the randomForest package is installed from CRAN.")
  }
  
  n.weights = rf$inbag
  n.col     = ncol(n.weights)
  N         = Matrix::Matrix(n.weights,sparse=TRUE)
  N.avg     = Matrix::Matrix(Matrix::rowMeans(N),nrow(N),1)
  
  pred      = CB_cforest(rf=rf,pb=prog.bar,rf.data=rf.data,pred.data=pred.data)$preds
  
  pred.aggr = rowMeans(pred)
  pred.cent = pred - pred.aggr
  pred.sums = Matrix::Matrix(rowSums(pred.cent),1,nrow(pred.cent))
  
  C   = N %*% t(pred.cent) - N.avg %*% pred.sums
  var = Matrix::colSums(C^2)/n.col^2
  
  return(data.frame('pred'=pred.aggr,var))
}

CB_cforest = function(rf,rf.data,pred.data=rf.data,pb=FALSE) 
{
  new.samples = match_sample(rf,rf.data)
  pb.fun      = ifelse(pb,pbapply::pblapply,lapply)
  out         = list()
  out$preds   = pb.fun(new.samples,function(x) honest.tree(formula=as.formula(rf$call$formula),data=x,pred.data=pred.data))
  out$preds   = do.call(cbind,out$preds)
  return(out)
}

match_sample <- function(rf,rf.data) {
  n = nrow(rf.data)
  matched.samples = lapply(1:ncol(rf$inbag),function(B){
                           n.uses            = rf$inbag[,B]
                           names(n.uses)     = 1:n
                           match.sample.rows = unlist(mapply(rep,x=as.numeric(names(n.uses)),each=n.uses))
                           matched.sample    = rf.data[match.sample.rows, ]
                           return(matched.sample)
                          })
  return(matched.samples)
}

honest.tree = function(formula0,data0,pred.data) 
{
  N      = dim(data0)[1]
  index  = sample(1:N, N/2)
  dat1   <<- data0[index,]
  dat2   = data0[-index,]
  result = rpart(formula0,data=dat1,control=rpart.control(minsplit=2,minbucket=1,cp=0.001))
  
  fit.party = as.party(result)
  
  pred.data$class.tree = predict(fit.party,newdata=pred.data,type="node")

  dat.pred = data.frame(class.tree = as.factor(predict(fit.party,newdata=dat2,type="node")),y0=as.numeric(dat2$y))
  agg.pred = aggregate(y0~class.tree,data=dat.pred,FUN=mean)
  
  if ( length(unique(result$where)) != length(unique(agg.pred$class.tree)) ) {
    index.unique = which( ! unique(result$where) %in% unique(agg.pred$class.tree))
    class.tree0  = unique(result$where)[index.unique]
    append       = data.frame(class.tree=as.factor(class.tree0),y0=rep(1/2,length(class.tree0)))
    agg.pred     = rbind(agg.pred,append)  
  }
  
  s.index              = sapply(pred.data$class.tree,function(x) which(x==agg.pred$class.tree))
  pred.data$pred       = agg.pred$y0[s.index]
  return(pred.data$pred)
}
```

```{r}
set.seed(333)
library(diffusionMap)
library(randomForest)
```

The following chunk loads the data analyzed in Freeman, Kim & Lee (2017): 2487 i-band (F814W) selected galaxies.
An Rdata file is a binary format; one creates Rdata files in R sessions by, for instance, doing the following...

save(variable1,variable2,...,variableN,file="foo.Rdata")

The variables can be any R variable type: data frames, matrices, vectors, etc.
```{r load i-band galaxy data}
load("statistics_iband.Rdata")
predictors = statistics.iband
```

In order to properly construct the diffusion map, one should standardize each relevant predictor (so that no
one predictor dominates object-to-object Euclidean distance measurements).
```{r standardize morphology statistics}
predictors.standardized = apply(predictors[,3:9], 2, scale)
rownames(predictors.standardized) = 1:nrow(predictors.standardized)
```

An implementation of diffusion map. See the Appendix of Freeman, Kim & Lee (2017) and references therein for
more detail. The relevant output is dmap$X--this is a matrix with 2487 rows and 30 columns. Only the first two
columns are used for visualization. In the context of the present analysis, this is sufficient to visualize
how high-mass- and low-mass-dominated regions lie in separate parts of the original predictor space.

NOTE: the orientation of the diffusion map is random, which is to say if you run the following code again and
again the map may flip along the x-axis as well as along the y-axis. So, at least initially, you may not get
a map whose orientation matches that of Figure 5 of Freeman, Kim & Lee (2017).
```{r implement diffusion map}
D       = as.matrix(dist(predictors.standardized))
nrow.D  = nrow(D)
K       = 30
sigma_i = rep(0,nrow.D)
for ( ii in 1:nrow.D ) {
  jj          = order(D[,ii])[K + 1]
  sigma_i[ii] = dist(rbind(predictors.standardized[ii,],predictors.standardized[jj,]))
}
S           = matrix(c(1/sqrt(sigma_i)),nrow.D,1) %*% matrix(c(1/sqrt(sigma_i)),1,nrow.D)
D           = D * S
dmap        = diffuse(D,eps.val=1)
plot(dmap$X[,1],dmap$X[,2],pch=15,col="grey",cex=0.75,xlab="First Diffusion Coordinate",ylab="Second Diffusion Coordinate",cex.axis=1.2,cex.lab=1.2)
```

Here we split the 2487 galaxies into training and test subsets (1787 and 700 galaxies each, respectively).
```{r data split}
set.seed(123123)
s                = sample(1:nrow(predictors),700)
predictors.test  = predictors[s,]
predictors.train = predictors[-s,]
```

Here we use quantile to determine the first and third quartiles (the 25th and 75th percentiles) of the mass.
We then split the data into three classes: H, M, and L (high, medium, low mass).
We create a data frame with the relevants columns of the predictor frame (3-9, for M, I, ..., A) and the response,
and remove the rows of the data frame for which the mass class is M. (Random forest is only applied to the upper
and lower quartile data.)
```{r mass training sample}
delta = 0.25
Q1 = quantile(predictors.train$mass,probs=delta)
Q3 = quantile(predictors.train$mass,probs=1-delta)

##### massClass indicates the name of mass groups - M: middle / H: high / L: Low
predictors.train$massClass <- "M"

for ( ii in 1:nrow(predictors.train) ) {
  if ( predictors.train$mass[ii] >= Q3) predictors.train$massClass[ii] <- "H"
  if ( predictors.train$mass[ii] <= Q1) predictors.train$massClass[ii] <- "L"
}

##### df.train contains summary statistics (col1-col7) and class label (col8) for "H" and "L" class objects
df.train.mass   = data.frame(predictors.train[,3:9],y=predictors.train[,ncol(predictors.train)])
df.train.mass   = df.train.mass[!df.train.mass$y=="M",]
df.train.mass$y = ifelse(df.train.mass$y=="H",1,0)
```

Here we define the test data frame.
```{r test sample}
df.test.mass = predictors.test[,3:9]
```

Here we run random forest along with the Wager et al./Wager & Athey-based tweaks.
alpha is the threshold p-value.
ratio is the percentage of high-mass galaxies (i.e., high/(high+low)). (The numerator in the ratio is the class that
is mapped to 1 above [see df.train.mass$y = ifelse...]. We map "H" to 1 in this analysis.) Here the ratio is ~1/2.
The outputs are mass.high, mass.low, and mass.notsig: this are the indices of *test-set* galaxies that are identified
as being in high-mass-dominated, low-mass-dominated, and "not significant" regions of the predictor space.
```{r RF for mass}
set.seed(1234)
alpha = 0.05
ratio = sum(df.train.mass$y)/length(df.train.mass$y)
rf      = try(randomForest(y~.,data=df.train.mass,keep.inbag=TRUE,sampsize=nrow(df.train.mass)^0.7,replace=FALSE,ntrees=1000))
ij.mass = rf_prediction(rf,rf.data=df.train.mass,pred.data=df.test.mass,prog.bar=FALSE)

mass.stat = (ij.mass$pred-ratio)/sqrt(ij.mass$var)
mass.pval = 2 * ifelse(pnorm(mass.stat,lower.tail = FALSE)< 1/2,pnorm(mass.stat,lower.tail=FALSE),pnorm(mass.stat,lower.tail=TRUE))
mass.pval = p.adjust(mass.pval,"BH") * sign(mass.stat)

imp = importance(rf)
dotchart(sort(imp[,1]),xlim=c(0,7),xlab="Mean Increase in Node Purity (Residual Sum of Squares)",pch=15,main="Mass Study",cex.axis=1.2,cex.lab=1.2)

high        = (mass.pval >= 0) & (mass.pval <  alpha) & (mass.stat > 0)
low         = (mass.pval <= 0) & (mass.pval > -alpha) & (mass.stat < 0)
mass.high   = which(high == T)
mass.low    = which(low == T)
mass.notsig = which(xor(high,low) == FALSE)
```

There are many ways one could proceed given mass.high, mass.low, and mass.notsig. One could construct boxplots, 
parallel coordinate plots, etc., etc. Here we plot the diffusion map coordinates for the test set data, highlighting
the high-mass-dominated region galaxies as red crosses ("red and dead"), and the low-mass-dominated region galaxies
as blue circles.
```{r plot diffusion map}
pch.plot = rep(15,length(s))
pch.plot[mass.high] = 4
pch.plot[mass.low] = 19
col.plot = rep("grey",length(s))
col.plot[mass.high] = "red"
col.plot[mass.low] = "blue"
cex.plot = rep(1.2,length(s))
cex.plot[c(mass.high,mass.low)] = 1.2
plot(dmap$X[s[mass.notsig],1],dmap$X[s[mass.notsig],2],pch=pch.plot[mass.notsig],cex=cex.plot[mass.notsig],col=col.plot[mass.notsig],main="",xlab="",ylab="",axes=FALSE)
points(dmap$X[s[mass.high],1],dmap$X[s[mass.high],2],pch=pch.plot[mass.high],cex=cex.plot[mass.high],col=col.plot[mass.high])
points(dmap$X[s[mass.low],1],dmap$X[s[mass.low],2],pch=pch.plot[mass.low],cex=cex.plot[mass.low],col=col.plot[mass.low])
```

