---
title: "Diffusion Map Visualization"
author: "Peter Freeman"
date: "May 18, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

HERE: read in predictor variables. Call this data frame "predictors"
```{r}
# Read in predictor variables.
```

Standardize the variables in each column of predictors.
```{r}
pred.stand           = apply(predictors,2,scale)
rownames(pred.stand) = 1:nrow(pred.stand)
```

Implement diffusion map in the manner of Freeman et al. (2017). As noted there, K=30 is a choice that works for our data, but
  there is no guarantee it will work best for *your* data. Play with this value.
Output: diffusion coordinates diff.coord, for all data. With good bookkeeping, you can keep track for which diffusion coordinates
  are those for training data, and which for test data.
```{r}
library(diffusionMap)

D       = as.matrix(dist(morph.stat))
nrow.D  = nrow(D)
K       = 30
sigma_i = rep(0,nrow.D)
for ( ii in 1:nrow.D ) {
  jj          = order(D[,ii])[K + 1]
  sigma_i[ii] = dist(rbind(morph.stat[ii,],morph.stat[jj,]))
}
S           = matrix(c(1/sqrt(sigma_i)),nrow.D,1) %*% matrix(c(1/sqrt(sigma_i)),1,nrow.D)
D           = D * S
dmap        = diffuse(D,eps.val=1)
diff.coord  = dmap$X

#plot(diff.coord[,1],diff.coord[,2],pch=19,cex=0.5,main="Diffusion Map",xlab="First Coordinate",ylab="Second Coordinate")
```